---
title: "Movement Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Collection and processing of data. 
```{r}
if(!dir.exists('./data')) {
    dir.create('./data')
}

fileUrl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
download.file(fileUrl, destfile = './data/data.zip')
unzip('./data/data.zip', exdir = './data')

activity <- read.csv("data/activity.csv")
str(activity)
```

## What is mean total number of steps taken per day?
```{r}
SumSteps <- tapply(activity$steps, activity$date, sum)
library(reshape2)
SumStepsMelted <- melt(SumSteps)
hist(SumStepsMelted$value, breaks = 20, xlab = "Total steps taken in a day", main = "", col = "orange")
```

### And the mean and median of the total steps in a day?
```{r}
mean(SumStepsMelted$value, na.rm = TRUE)
median(SumStepsMelted$value, na.rm = TRUE)
```

## What is the average daily activity pattern?
### Time series plot of 5 min intervals and average number of steps taken. 
```{r}
ActAvg <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
ActAvgMelt <- melt(ActAvg)
names(ActAvgMelt) <- c("Interval", "AvgSteps")
plot(AvgSteps ~ Interval, data = ActAvgMelt, type = "l", col = "red", main = "Averge daily steps", xlab = "5 min interval")
```

### And the mean and median total number of steps in a day? 
```{r}
ActAvgMelt[ActAvgMelt$AvgSteps == max(ActAvgMelt$AvgSteps), ]
```
### So the maximal mean daily steps occurs at the 835th 5 minute interval


## There are days and intervals where data are missing. This could affect the analysis. 
```{r}
sum(is.na(activity$steps))/nrow(activity)
```
### So 13% of the observations are not actually measured. 

## Impute missing values
### Create new data set with imputed missing values, using mean without NAs, for comparison. 
```{r}
ActImpute <- activity
ActImpute$steps[is.na(ActImpute$steps)] <- mean(ActImpute$steps, na.rm = TRUE)
```
### Make a histogram of the total number of steps taken each day.
```{r}
ActImputeSteps <- tapply(ActImpute$steps, ActImpute$date, sum)
library(reshape2)
ImputeMelt <- melt(ActImputeSteps)
names(ImputeMelt) <- c("Date", "SumOfSteps")
MeanSteps <- mean(ImputeMelt$SumOfSteps)
MedianSteps <- median(ImputeMelt$SumOfSteps)
hist(ImputeMelt$SumOfSteps, breaks = 20, xlab = "Total steps taken in a day", main = "", col = "red")
```


### Report mean and median of total number of steps taken per day.
```{r}
mean(ImputeMelt$SumOfSteps, na.rm = TRUE)
median(ImputeMelt$SumOfSteps, na.rm = TRUE)
```

### Compare these values to when the NAs were just removed, above: mean = 10766.19 and median = 10765. Not a great deal of difference. 

## Are there differences in activity patterns between weekdays and weekends? 
### Looking only at the imputed data create a new column which tells whether a date is a weekday or weekend day.
```{r}
ActImpute$date <- as.Date(ActImpute$date)
ActImpute$weekdays <- weekdays(ActImpute$date)
ActImpute$weeks[(ActImpute$weekdays == "Saturday" | ActImpute$weekdays == "Sunday")] <- "weekend"
ActImpute$weeks[!(ActImpute$weekdays == "Saturday" | ActImpute$weekdays == "Sunday")] <- "weekdays"
library(plyr)
WeekData <- ddply(ActImpute, c("interval", "weeks"), function(x) apply(x[1], 2, mean))
str(WeekData)
```
### Panel plot of time series of 5-minute intervals and average steps taken, averaged across all weekday days or weekend days.
```{r}
library(lattice)
xyplot(steps ~ interval | weeks, data = WeekData, type = "l", xlab = "interval", ylab = "steps", layout = c(1, 2))
```


### Notice activity is spread across the day during the weekend and concentrated roughly before midday during the week. 